---
title: Modelling Land-use Impacts on Biodiversity with the PREDICTS Database
subtitle: December 2024 Update to Include Bayesian Models
date: "`r Sys.Date()`"
author: "Tim Newbold"
output:
  rmdformats::downcute:
    code_folding: show
    self_contained: true
    thumbnails: false
    lightbox: true
references:
  - id: Hudson2017
    title: The database of the PREDICTS (Projecting Responses of Ecological Diversity In Changing Terrestrial Systems) project
    author:
      - family: Hudson
        given: Lawrence N.
      - family: Newbold
        given: Tim
      - family: Contu
        given: Sara
      - family: Hill
        given: Samantha L. L.
      - family: Lysenko
        given: Igor
      - family: De Palma
        given: Adriana
      - family: Phillips
        given: Helen R. P.
      - family: Data entrants
      - family: Data contributors
      - family: Collen
        given: Ben
      - family: Ewers
        given: Rob M.
      - family: Mace
        given: Georgina M.
      - family: Purves
        given: Drew W.
      - family: Scharlemann
        given: J&oumlaut;rn P. W.
      - family: Purvis
        given: Andy
    container-title: Ecology & Evolution
    volume: 7
    page: 145-188
    DOI: 10.1002/ece3.2579
    URL: https://doi.org/10.1002/ece3.2579
    type: article-journal
    issued:
      year: 2017
  - id: Ladle2011
    title: Conservation Biogeography
    author:
      - family: Ladle
        given: Richard J.
      - family: Whittaker
        given: Robert J.
    type: book
    publisher: Wiley-Blackwell, Chichester, UK
    issued:
      year: 2011
  - id: Olson2001
    title: Terrestrial ecoregions of the world&colon; a new map of life on Earth
    author:
      - family: Olson
        given: David M.
      - family: Dinerstein
        given: Eric
      - family: Wikramanayake
        given: Eric D.
      - family: Burgess
        given: Neil D.
      - family: Powell
        given: George V.N.
      - family: Underwood
        given: Emma C.
      - family: D'Amico
        given: Jennifer A.
      - family: Itoua
        given: Illanga
      - family: Strand
        given: Holly E.
      - family: Morrison
        given: John C.
      - family: Loucks
        given: Colby J.
      - family: Allnutt
        given: Thomas F.
      - family: Ricketts
        given: Taylor
      - family: Kura
        given: Yumiko
      - family: Lamoreux
        given: John F.
      - family: Wettengel
        given: Wesley W.
      - family: Hedao
        given: Prashant
      - family: Kassem
        given: Kenneth R.
    container-title: Bioscience
    volume: 51
    page: 933-938
    URL: https://academic.oup.com/bioscience/article/51/11/933/227116
    type: article-journal
    issued:
      year: 2001
  - id: Myers2000
    title: Biodiversity hotspots for conservation priorities
    author:
      - family: Myers
        given: Norman
      - family: Mittermeier
        given: Russell A.
      - family: Mittermeier
        given: Cristina G.
      - family: da Fonseca
        given: Gustavo A.B.
      - family: Kent
        given: Jennifer
    container-title: Nature
    volume: 403
    page: 853-858
    type: article-journal
    issued:
      year: 2000
  - id: Brooks2006
    title: Global biodiversity conservation priorities
    author:
      - family: Brooks
        given: Thomas M.
      - family: Mittermeier
        given: Russell A.
      - family: da Fonseca
        given: Gustavo A.B.
      - family: Gerlach
        given: J.
      - family: Hoffmann
        given: Michael
      - family: Lamoreux
        given: John F.
      - family: Mittermeier
        given: Cristina G.
      - family: Pilgrim
        given: John D.
      - family: Rodrigues
        given: Ana S.L.
    container-title: Science
    volume: 313
    page: 58-61
    DOI: 10.1126/science.1127609
    URL: https://doi.org/10.1126/science.1127609
    type: article-journal
    issued:
      year: 2006
---

```{r loadpackages, echo=FALSE, eval=TRUE, warning=FALSE, message=FALSE}
library(kableExtra)
library(ggplot2)
library(ggwordcloud)
library(dplyr)
library(tidyverse)
library(tidyr)
library(stringr)
library(terra)
library(tidyterra)
library(treemapify)
```

# Introduction

The PREDICTS database [@Hudson2017] is a fully open-access database containing spatial comparisons of biodiversity in different land uses worldwide. 

You can get started quickly by downloading the PREDICTS database from my website, but if you intend to use the database in your research, I would ask you please to download the database (freely) from the <a href="http://doi.org/10.5519/0066354" target="_blank">Natural History Museum data portal</a>, so that appropriate credit is given.

```{r readdatabase,echo=TRUE,results=TRUE}
predicts <- url("https://timnewbold.github.io/predicts_database.rds?dl=1") %>% readRDS()
```

To work with the database, you should install a package that contains some basic data-manipulation functions:

```{r installpredictsfunctions,echo=TRUE,results=FALSE,eval=FALSE}
remotes::install_github("timnewbold/predicts-demo",subdir="predictsFunctions")
```

```{r loadpredictsfunctions,echo=TRUE,results=TRUE}
library(predictsFunctions)
```

# Database Structure

```{r structurefields,echo=FALSE,eval=TRUE}
colsStruct <- data.frame(Field=c('Source_ID',
                           'Study_number',
                           'Study_name',
                           'SS',
                           'Block',
                           'SSB',
                           'Site_number',
                           'Site_name',
                           'SSBS',
                           'SSS',
                           'Measurement'),
                   Description=c('The published paper (***Source***) from which data were derived',
                                 'Within each ***Source***, the ***Study*** identifier',
                                 'A text description to identify the ***Study***',
                                 'An amalgamation of ***Source*** and ***Study*** IDs, so that each ***Study*** in the database is uniquely identifiable',
                                 'A text description of distinct spatial ***Blocks***, blank where sites were not spatially clustered',
                                 'An almagamation of the unique ***Study*** identifier, *SS*, and *Block*, so that each ***Block*** is uniquely identifiable',
                                 'Within each ***Study***, the ***Site*** identifier',
                                 'A text description to identify the ***Site***',
                                 'An almagamation of the unique ***Block*** identifier, *SSB*, and the ***Site*** number, so that each ***Site*** is uniquely identifiable',
                                 'As for *SSBS*, but excluding the block identifier - not recommended, but can be used if you don\'t want to account for spatial structure',
                                 'The recorded ***Measurement*** of species abundance, species occurrence, or species richness (see next section)'))
```

The PREDICTS database has a hierarchical structure.

At the highest level, the data in the 2016 release of the PREDICTS database are drawn from `r format(n_distinct(predicts$Source_ID),big.mark=",")` published papers (***Sources***) comparing biodiversity in different land uses. 

Each data Source consists of one or more ***Studies***. Sometimes, a published paper will contain biodiversity samples collected using more than one fundamental approach. For example, a paper may contain data for different taxonomic groups that would not appropriately have been sampled using the same methods. In such cases, because data are not directly comparable if collected using different methods, data Sources were split into multiple Studies. Otherwise, each Source contained a single Study.

Each Study consists of data sampled in one or more spatial ***Blocks***. A Study was divided into more than one Block if the sampling showed spatial structuring, to allow us to delimit spatial aggregations of Sites, and thus to account for spatial structuring later when modelling the data.

Each ***Block*** contains two or more ***Sites***, which are the specific locations at which biodiversity was sampled.

The complete 2016 version of the PREDICTS database contains data from `r format(n_distinct(predicts$SS),big.mark=",")` ***Studies***, which together sample a total of `r format(n_distinct(predicts$SSBS),big.mark=",")` ***Sites***.

Within each ***Site***, there are a series of ***Measurements*** - of species abundance (`r format(table(predicts$Diversity_metric_type)["Abundance"],big.mark=",")` records), species occurrence (`r format(table(predicts$Diversity_metric_type)["Occurrence"],big.mark=",")` records), or species richness (`r format(table(predicts$Diversity_metric_type)["Species richness"],big.mark=",")` records) - for each of the sampled taxa.

```{r figdiversitymetrics, echo=FALSE,eval=TRUE, fig.cap="Tree map showing the frequency of different biodiversity metric types across the PREDICTS database. Note that counts of metrics were square-root transformed to counter the strong dominance of abundance measures"}

MetricCounts <- predicts %>% count(Diversity_metric_type) %>%
  mutate(Diversity_metric = str_to_title(Diversity_metric_type))

p <- ggplot(data = MetricCounts,
            mapping = aes(area = sqrt(n),
                          fill = Diversity_metric_type,
                          label = Diversity_metric_type)) + 
  geom_treemap() + 
  guides(fill = guide_legend(title = "Diversity Metric Type"))
  # geom_treemap_text(colour = "#000000", grow=TRUE,min.size = 1,reflow = TRUE)# + 
  # theme(legend.position = "none")

p

```

```{r structuretable,echo=FALSE,eval=TRUE}
knitr::kable(colsStruct,caption="Key fields in the PREDICTS database relating to the hierarchical structure of the database")
```

# Key Sampling Information

```{r samplingfields,echo=FALSE,eval=TRUE}
colsSampling <- data.frame(Field=c('Measurement',
                                   'Diversity_metric_type',
                                   'Diversity_metric',
                                   'Diversity_metric_unit',
                                   'Diversity_metric_is_effort_sensitive',
                                   'Sampling_method',
                                   'Sampling_effort',
                                   'Rescaled_sampling_effort',
                                   'Sampling_effort_unit',
                                   'Sample_start_earliest',
                                   'Sample_end_latest',
                                   'Sample_date_resolution',
                                   'Sample_midpoint',
                                   'Max_linear_extent_metres'),
                   Description=c('The recorded ***Measurement*** of species abundance, species occurrence, or species richness',
                                 'Whether the recorded ***Measurement*** is of species abundance, species occurrence or species richness',
                                 'The specific metric used for the taxon ***Measurement***',
                                 'The units of the diversity metric',
                                 'Whether the diversity metric is sensitive to sampling effort',
                                 'The method used to sample taxa',
                                 'The sampling effort applied at the site',
                                 'Rescaled sampling effort, where the maximum within a study becomes 1, and everything else a fraction of the maximum',
                                 'The units of the sampling effort',
                                 'The earliest date at which sampling may have begun',
                                 'The latest date at which sampling may have ended',
                                 'The resolution of the sampling start and end dates (day, month or year)',
                                 'The mid-point between earliest sampling start and latest sampling end',
                                 'The maximum linear span sampled at a given site'))
```


Measurements in PREDICTS consist of three broad types: species abundance, species occurrence (i.e., presence or absence), or species richness (the number of species sampled). Within these categories, there are lots of different specific metrics. Some measures of diversity are inherently sensitive to sampling effort, which has important implications for processing the data (see below).

```{r figmetrictypes, echo=FALSE,eval=TRUE, fig.cap="Tree map showing the frequency of different biodiversity metric types across the PREDICTS database. Note that counts of metrics were square-root transformed to counter the strong dominance of abundance measures"}

MetricCounts <- predicts %>% count(Diversity_metric_type) %>%
  mutate(Diversity_metric = str_to_title(Diversity_metric_type))

p <- ggplot(data = MetricCounts,
            mapping = aes(area = sqrt(n),
                          fill = Diversity_metric_type,
                          label = Diversity_metric_type)) + 
  geom_treemap() + 
  guides(fill = guide_legend(title = "Diversity Metric Type"))
  # geom_treemap_text(colour = "#000000", grow=TRUE,min.size = 1,reflow = TRUE)# + 
  # theme(legend.position = "none")

p

```

The different underlying ***Studies*** in PREDICTS consist of data sampled using a whole host of different methods, and with widely differing levels of sampling effort. When analysing the PREDICTS data, it is important that diversity ***Measurements*** are comparable within ***Studies***, whereas differences among ***Studies*** are handled by the hierarchical design of the models (see below). Therefore, it is a requirement of the PREDICTS database that sampling *methods* are constant within studies. Nevertheless, for `r round((length(which(tapply(X = predicts$Sampling_effort,INDEX = predicts$SS,FUN = function(s) return(diff(range(s))))>0))/length(unique(predicts$SS)))*100,0)`% of studies, sampling effort differs among the sampled sites. We will see how we deal with this later.

Most of the original biodiversity sampling in the studies contained within the PREDICTS database took place between 2000 and 2013.

```{r figsampleyears,echo=FALSE,eval=TRUE,fig.cap="Temporal density of mid-point sampling dates for the PREDICTS sites",message=FALSE,warning=FALSE,fig.width=12,fig.height=8}

sites <- predicts %>%
  select(c('SSS','Sample_midpoint','Max_linear_extent_metres',
           'Latitude','Longitude','Biome')) %>%
  distinct() %>% 
  mutate(Year = as.numeric(format(
    as.POSIXct(Sample_midpoint),format="%Y")))

p <- ggplot() + 
  geom_density(data = sites,mapping = aes(x = Year), col = "#2F67B1",fill = "#2F67B133") + 
  scale_y_continuous(name = "Site Density") + 
  theme_classic(base_size = 25)

p

```

Sampling at the sites in PREDICTS can span a range of different spatial extents. The maximum linear span of sampling for three-quarters of sites in the PREDICTS database falls between `r round(quantile(sites$Max_linear_extent_metres,probs=0.125,na.rm=TRUE),0)` and `r round(quantile(sites$Max_linear_extent_metres,probs=0.875,na.rm=TRUE),0)` metres.

```{r figsampleextents,echo=FALSE,eval=TRUE,fig.cap="Density distribution of maximum linear spans sampled at the PREDICTS sites. The error bar represents the central 75% of values in the distribution",message=FALSE,warning=FALSE,fig.width=12,fig.height=8}

sites2 <- sites %>%
  subset(!is.na(Max_linear_extent_metres))

MLE_quantiles <- sites %>% 
  summarise(q = quantile(
    Max_linear_extent_metres,probs = c(0.125,0.875),
    na.rm=TRUE))

p <- ggplot() + 
  geom_density(data = sites2,mapping = aes(x = Max_linear_extent_metres),
               col = "#2F67B1",fill = "#2F67B133") + 
  geom_errorbarh(mapping = aes(y = 1, xmin = MLE_quantiles$q[1],
                               xmax = MLE_quantiles$q[2], height = 0.02), 
                 size = 1.5,col = "#1065AB") +
  scale_x_continuous(name = "Maximum linear extent (m)",transform = 'log10',
                     breaks = c(0.1,1,10,100,1000,10000),labels = scales::comma) +
  scale_y_continuous(name = "Site Density") +
  theme_classic(base_size = 25)

p

```

```{r samplingtable,echo=FALSE,eval=TRUE}
knitr::kable(colsSampling,caption="Key fields in the PREDICTS database describing biodiversity sampling")
```

# Spatial Information

```{r spatialfields,echo=FALSE,eval=TRUE}
colsSpatial <- data.frame(Field=c('Longitude',
                                  'Latitude',
                                  'Realm',
                                  'Biome',
                                  'Ecoregion',
                                  'UN_region',
                                  'UN_subregion',
                                  'Hotspot',
                                  'Wilderness_area'),
                   Description=c('The longitude of the site at which sampling was undertaken',
                                           'The latitude of the site at which sampling was undertaken',
                                           'Based on the coordinates, the location of the site within...biogeographic realms',
                                 '...biomes',
                                 '...ecoregions',
                                 '...UN regions',
                                 '...UN sub-regions',
                                 '...Biodiversity hotspots',
                                 '...High-biodiversity wilderness areas'))
```

The vast majority of sampled sites in the PREDICTS database (`r round((length(which(!is.na(sites$Longitude)))/nrow(sites))*100,1)`%) are associated with geographical coordinates (longitude and latitude).

These geographical coordinates are then used to assign the site within various geographic schemes: biogeographic realms [@Ladle2011; @Olson2001], biomes [@Olson2001], ecoregions [@Olson2001], UN regions and sub-regions (<a href="https://unstats.un.org/unsd/methodology/m49/" target="_blank">United Nations, 1999</a>), biodiversity hotspots [@Myers2000] and high-biodiversity wilderness areas [@Brooks2006].

```{r, echo=FALSE,eval=TRUE, warning=FALSE}

studies <- predicts %>% group_by(SS) %>% 
  summarise(n_spp = n_distinct(Taxon_name_entered),
            Longitude = mean(Longitude),
            Latitude = mean(Latitude))

n.studies <- nrow(studies)
n.sites <- nrow(sites)

```

The sites in PREDICTS are distributed across all of the world's 14 terrestrial biomes, `r n_distinct(predicts$Ecoregion)` ecoregions, and `r n_distinct(predicts$Country)` countries.

```{r figsitesmap, echo=FALSE,eval=TRUE, warning=FALSE, fig.cap="Map of sites with biodiversity samples in the PREDICTS database"}

world <- "C:/Users/tim_n/Downloads/outline_clip2.shp" %>% 
  vect()#  %>% geom_sf() %>% fortify()

p <- ggplot() + 
  geom_spatvector(data = world) + 
  geom_point(data = studies,
             mapping = aes(x = Longitude,y = Latitude,size=n_spp),
             col = "#2E258566") +
  scale_x_continuous(name = "") + 
  scale_y_continuous(name = "") + 
  theme_classic()

p
```

```{r spatialtable,echo=FALSE,eval=TRUE}
knitr::kable(colsSpatial,caption="Key fields in the PREDICTS database describing the spatial location of biodiversity sampling")
```

# Human Pressure Information

```{r,echo=FALSE,eval=TRUE}
colsPressure <- data.frame(Field=c('Habitat_as_described',
                                  'Predominant_land_use',
                                  'Source_for_predominant_land_use',
                                  'Use_intensity'),
                   Description=c('The description of the habitat from the original source publication',
                                           'The land-use type, as classified by the data entrant',
                                           'Whether land-use type was classified based on information given in the original source publication or Google Maps',
                                 'Intensity of human land use, as classified by the data entrant'))
```

The only information on human pressures that is endogenous to the PREDICTS database describes the land-use conditions at the sampled site, with the key variables representing land-use type and intensity of human land use.

In the majority of cases (at least `r round((length(which(predicts$Source_for_predominant_land_use=="Direct from publication / author"))/nrow(predicts))*100,1)`%), land-use variables in the PREDICTS database were encoded using information given in the original source publications. Occasionally, the data entrant used Google Maps to infer the land-use type.

Land-use type is classified as one of: primary vegetation (apparently pristine habitat, with no record of prior destruction); secondary vegetation (habitat previously destroyed by human actions or extreme natural events, but now recovering towards its natural state); plantation forest (area used for growing woody crops); cropland (area used for growing herbaceous crops); pasture (area used for grazing livestock); urban (human settlements or areas of civic amenity). Secondary vegetation is also divided according to the stage of recovery toward its natural state: young, intermediate or mature. For full definitions, see [@Hudson2017].

Land-use intensity is classified at one of three levels (minimal, light or intense), with the criteria for classification depending on land-use type (see [@Hudson2017] for more details)

```{r figlanduses, echo=FALSE,eval=TRUE, warning=FALSE,fig.cap="Distribution of sites among major land uses (note that all stages of recovery of secondary vegetation have been combined in this figure"}

sites3 <- predicts %>%
  select(c('SSS','Latitude','Longitude','Predominant_land_use')) %>%
  mutate(LandUse = recode(Predominant_land_use,
                          'Mature secondary vegetation' = 'Secondary vegetation',
                          'Intermediate secondary vegetation' = 'Secondary vegetation',
                          'Young secondary vegetation' = 'Secondary vegetation',
                          'Secondary vegetation (indeterminate age)' = 'Secondary vegetation',
                          'Cannot decide' = NA_character_)) %>%
  mutate(LandUse = fct_relevel(LandUse,
                               c("Primary vegetation","Secondary vegetation",
                                 "Plantation forest","Cropland","Pasture","Urban"))) %>%
  distinct()

LUCounts <- sites3 %>% count(LandUse) %>%
  mutate(LandUse = str_to_title(LandUse)) %>%
  mutate(LandUse = factor(LandUse,
                          levels=c("Primary Vegetation","Secondary Vegetation",
                                   "Plantation Forest","Cropland",
                                   "Pasture","Urban")))

p <- ggplot(data = LUCounts,
            mapping = aes(area = n,
                          fill = LandUse,
                          label = LandUse)) + 
  geom_treemap() + 
  guides(fill = guide_legend(title = "Land Use"))

p

```

```{r tablepressure,echo=FALSE,eval=TRUE}
knitr::kable(colsPressure,caption="Key fields in the PREDICTS database describing human land-use pressures")
```

# Taxonomic Information

```{r taxonomicfields,echo=FALSE,eval=TRUE}
colsTax <- data.frame(Field=c('Taxon_name_entered',
                              'Parsed_name',
                              'Taxon',
                              'Name_status',
                              'Rank',
                              'Kingdom',
                              'Phylum',
                              'Class',
                              'Order',
                              'Family',
                              'Genus',
                              'Species',
                              'Best_guess_binomial',
                              'Study_common_taxon'),
                   Description=c('The name of the sampled taxon, as given in the original source publication',
                                 'The name of the sampled taxon, after resolving any typos',
                                 'The accepted or provisionally accepted name for the species',
                                 'Whether the taxon name is accepted or provisionally accepted',
                                 'The taxonomic rank of the accepted/provisionally accepted taxon name',
                                 'For the accepted/provisionally accepted taxon name, the name of the ...Kingdom',
                                 '...Phylum',
                                 '...Class',
                                 '...Order',
                                 '...Family',
                                 '...Genus',
                                 '...Species',
                                 'For as many sampled taxa as possible, the best guess of the Latin binomial name',
                                 'The taxonomic group within which all taxa sampled in a particular study belong'))
```

The names of the sampled taxa, as given in the original source publications, are entered directly into the database. These names were passed through a taxonomic names resolver to attempt to correct any typographical errors from the original papers. Parsed names were then compared to the Catalogue of Life or manually checked, to try to find the accepted name for the species.

A total of `r format(n_distinct(predicts$Taxon_name_entered),big.mark=",")` taxonomic names are given in the original publications that underpin the PREDICTS database. After taxonomic resolution, a total of `r format(n_distinct(predicts$Best_guess_binomial),big.mark=",")` accepted Latin binomials are identified (many taxa are identified to a higher level than species, or are identified only as morpho-species).

```{r figtaxa, echo=FALSE,eval=TRUE, warning=FALSE,fig.cap="Tree map representing the number of species sampled in different phyla"}

TaxonCounts <- predicts %>% 
  subset(Phylum!="") %>%
  select(Kingdom,Phylum,Class,Order,Family,Best_guess_binomial) %>% 
  distinct() %>%
  count(Phylum)

p <- ggplot(data = TaxonCounts,
            mapping = aes(area = n,
                          fill = Phylum,
                          label = Phylum)) + 
  geom_treemap() + 
  geom_treemap_text(colour = "#000000", grow=TRUE,min.size = 10,reflow = TRUE) + 
  guides(fill = guide_legend(title = "Phylum"))

p

```

```{r tabletaxonomic,echo=FALSE,eval=TRUE}
knitr::kable(colsTax,caption="Key fields in the PREDICTS database relating to the taxonomy of sampled taxa")
```

# Data Pre-processing

Before modelling with the PREDICTS data, there are a few basic pre-processing steps that must be carried out. First, for the `r round((as.integer(summary(predicts$Measurement==predicts$Effort_corrected_measurement)['FALSE'][[1]])/nrow(predicts))*100,1)`% of records that are of an effort-sensitive metric, and where sampling effort differs among sampled sites within the study, we need to correct sampling effort. This is done by rescaling sampling effort within each study, such that the most-sampled site gets a relative effort value of 1, and the sampling effort for all other sites is scaled linearly relative to this maximum value. The _CorrectSamplingEffort_ function in the _predictsFunctions_ package carries out these operations:

```{r,echo=TRUE,results=TRUE}
# Correct effort-sensitive abundance measures (assumes linear relationship between effort and recorded abundance)
predicts <- predicts %>% CorrectSamplingEffort()
```

Next, some samples in PREDICTS were entered such that a sampled 'Site' consisted of very finely divided samples. For example, samples may have been entered as individual traps. This is not in keeping with the strict definition of a 'Site' in the PREDICTS database. To get around this issue, we combine sites with identical coordinates, belonging to the same study and spatial block of sites, sampled on the same dates, sampled with the same methods and recorded using the same diversity metric, and situated within the same land-use type and human land-use intensity. This merging of sites is carried out by the _MergeSites_ function in the _predictsFunctions_ package (warning, this operation can take a few minutes to run, so you might want to go and make a cup of tea!):

```{r,echo=TRUE,results=TRUE}
predicts <- predicts %>% MergeSites(silent=TRUE)
```

There are two main ways that you can now analyze the PREDICTS data:

1. Using the raw species-level data, modelling the presence/absence or abundance of individual species; or
2. Using site-level biodiversity summaries, such as species richness, total community abundance or other aggregate biodiversity measures

We are now ready to build some simple models using the PREDICTS database. For the modelling, we will use the brms package.

```{r,echo=TRUE,results=TRUE,message=FALSE,warning=FALSE}
library(brms)
```

# Basic Species-level Models

One way we can use the PREDICTS data is to build models based on the raw data for individual species. 

So that the models run in a reasonably short amount of time, we will focus here on data for nymphalid butterflies. We need to select records only of abundance or occurrence. We will then create a variable called "*Occurence*" that represents the presence or absence of the species. Before we build a model, we need to rearrange the land-use classification a bit. Any sites where the land-use type was entered as 'Cannot decide' need to be set as NA values. We will also group together secondary vegetation sites at all stages of recovery, and remove sites in urban areas, owing to the relatively limited number of sites for nymphalid butterflies in these land-use types.

```{r,echo=TRUE,results=TRUE}
nymphals <- predicts %>%
  subset(Family == "Nymphalidae") %>% 
  subset(Diversity_metric_type %in% c("Abundance","Occurrence")) %>%
  mutate(Occurrence = ifelse(Measurement>0,1,0)) %>%
  mutate(LandUse = recode(Predominant_land_use,
    'Primary vegetation' = 'Primary',
                                 'Young secondary vegetation' = 'Secondary',
                                 'Intermediate secondary vegetation' = 'Secondary',
                                 'Mature secondary vegetation' = 'Secondary',
                                 'Secondary vegetation (indeterminate age)' = "Secondary",
                                 'Plantation forest' = 'Plantation',
                                 'Cropland' = 'Cropland',
                                 'Pasture' = 'Pasture',
                                 'Urban' = 'Urban',
                                 'Cannot decide' = NA_character_)) %>%
  subset(LandUse != "Urban")
  droplevels()
```

Let's try a very simple model fitting just land use.

```{r,echo=TRUE,results=TRUE}
occMod <- nymphals %>% brm(formula = Occurrence ~ 
                          LandUse+(1|SS)+(1|SSBS)+(1|Taxon_name_entered), 
                        family = "bernoulli",iter = 2000,chains = 4,cores = 4)




```

If we look at the pseudo R<sup>2</sup> values of the model, we can see that the random effects explain most of the variation in reptile species presence/absence, while land use explains `r round(StatisticalModels::R2GLMER(occMod1$model)$marginal*100,1)`%.

```{r,echo=TRUE,results=TRUE}
StatisticalModels::R2GLMER(occMod1$model)
```

Plotting this model shows significant declines in the probability of occurrence of reptile species in secondary vegetation, plantation forests and pastures.

```{r,echo=TRUE,results=TRUE}
PlotGLMERFactor(model = occMod1$model,data = occMod1$data,responseVar = "Prob. occurrence",logLink = "b",catEffects = "LandUse")
```
