---
title: Modelling Land-use Impacts on Biodiversity with the PREDICTS Database
date: "`r Sys.Date()`"
author: "Tim Newbold"
output:
  rmdformats::downcute:
    code_folding: show
    self_contained: true
    thumbnails: false
    lightbox: true
references:
  - id: Hudson2017
    title: The database of the PREDICTS (Projecting Responses of Ecological Diversity In Changing Terrestrial Systems) project
    author:
      - family: Hudson
        given: Lawrence N.
      - family: Newbold
        given: Tim
      - family: Contu
        given: Sara
      - family: Hill
        given: Samantha L. L.
      - family: Lysenko
        given: Igor
      - family: De Palma
        given: Adriana
      - family: Phillips
        given: Helen R. P.
      - family: Data entrants
      - family: Data contributors
      - family: Collen
        given: Ben
      - family: Ewers
        given: Rob M.
      - family: Mace
        given: Georgina M.
      - family: Purves
        given: Drew W.
      - family: Scharlemann
        given: J&oumlaut;rn P. W.
      - family: Purvis
        given: Andy
    container-title: Ecology & Evolution
    volume: 7
    page: 145-188
    DOI: 10.1002/ece3.2579
    URL: https://doi.org/10.1002/ece3.2579
    type: article-journal
    issued:
      year: 2017
---

# Introduction

The PREDICTS database [@Hudson2017] is a fully open-access database containing spatial comparisons of biodiversity in different land uses worldwide. 

You can get started quickly by downloading the PREDICTS database from my website, but if you intend to use the database in your research, I would ask you please to download the database (freely) from the <a href="http://doi.org/10.5519/0066354" target="_blank">Natural History Museum data portal</a>, so that appropriate credit is given.

```{r,echo=TRUE,results=TRUE}
webFile <- url("https://timnewbold.github.io/predicts_database.rds?dl=1")
predicts <- readRDS(webFile)
```

To work with the database, you should install a package that contains some basic data-manipulation functions:

```{r,echo=TRUE,results=FALSE,eval=FALSE}
remotes::install_github("timnewbold/predicts-demo",subdir="predictsFunctions")
```

```{r,echo=TRUE,results=TRUE}
library(predictsFunctions)
```

# Database Structure

The PREDICTS database consists of data drawn from `r length(unique(predicts$Source_ID))` published papers comparing biodiversity in different land uses.

# References