---
title: Multi-model Selection
date: "`r Sys.Date()`"
author: "Tim Newbold"
output:
  rmdformats::downcute:
    code_folding: show
    self_contained: true
    thumbnails: false
    lightbox: true
references:
  - id: Hudson2017
    title: The database of the PREDICTS (Projecting Responses of Ecological Diversity In Changing Terrestrial Systems) project
    author:
      - family: Hudson
        given: Lawrence N.
      - family: Newbold
        given: Tim
      - family: Contu
        given: Sara
      - family: Hill
        given: Samantha L. L.
      - family: Lysenko
        given: Igor
      - family: De Palma
        given: Adriana
      - family: Phillips
        given: Helen R. P.
      - family: Data entrants
      - family: Data contributors
      - family: Collen
        given: Ben
      - family: Ewers
        given: Rob M.
      - family: Mace
        given: Georgina M.
      - family: Purves
        given: Drew W.
      - family: Scharlemann
        given: J&oumlaut;rn P. W.
      - family: Purvis
        given: Andy
    container-title: Ecology & Evolution
    volume: 7
    page: 145-188
    DOI: 10.1002/ece3.2579
    URL: https://doi.org/10.1002/ece3.2579
    type: article-journal
    issued:
      year: 2017
  - id: Gould2013
    title: Forest restoration and parasitoid wasp communities in montane Hawai'i
    author:
      - family: Gould
        given: Rachelle K.
      - family: Pejchar
        given: Liba
      - family: Bothwell
        given: Sara G.
      - family: Brosi
        given: Berry
      - family: Wolny
        given: Stacie
      - family: Mendenhall
        given: Chase
      - family: Daily
        given: Gretchen
    container-title: PLoS ONE
    volume: 8
    page: e59356
    DOI: 10.1371/journal.pone.0059356
    URL: https://doi.org/10.1371/journal.pone.0059356
    type: article-journal
    issued:
      year: 2013
  - id: Burnham2002
    title: Model Selection and Multimodel Inference&colon; An Information Theoretic Approach
    author:
      - family: Burnham
        given: K.P.
      - family: Anderson
        given: D.R.
    type: book
    publisher: Springer-Verlag, New York
    edition: 2nd
    issued:
      year: 2002
---

# Overview

In this session, I will introduce multi-model selection as an alternative to backward stepwise model selection. Classical frequentist statistics is based on P values, which give the probability of obtaining an observed dataset given the null hypothesis of no relationship between explanatory and response variables. If the P value is below a certain threshold, $\alpha$ (often 0.05), then an explanatory variable is deemed to have a significant effect on the response variable. Explanatory variables with a P value above this threshold are deemed not to have a significant effect, and are therefore discarded. The problem with this approach is that the P value is rather arbitrary, and the quest to find significant results can lead to some rather dubious behaviours (collectively often referred to as 'P hacking').

An alternative approach, so-called 'multi-model selection' was proposed by [@Burnham2002]. The idea of this approach is to assess the relative support for each model in a candidate set of models. Models are compared using Information Criteria, the most widely used of which is the Akaike Information Criterion (AIC). 

# Akaike Information Criterion (AIC)

Information criteria measure how well a model fits the data (measured using the likelihood of the model), with some penalty applied for model complexity. The AIC is calculated as:
$$AIC=-2\times LL_{model}+2\times DF$$

Where $LL_{model}$ is the log-likelihood of the model, and $DF$ is the model degrees of freedom.

As in the last session, you will need my _MResModelling_ R package:

```{r,echo=TRUE,results=FALSE,eval=FALSE}
remotes::install_github("timnewbold/MResEcologicalModelling",subdir="MResModelling")
```

```{r,echo=TRUE,results=TRUE,warning=FALSE,message=FALSE}
library(MResModelling)
```

# Example Dataset

We will work again with the dataset describing the abundance of the parasitoid wasp species *Pimpla punicipes* in sites of different land use in Hawai'i.

```{r,echo=TRUE,results=TRUE}
# Load species-level data (loads as object 'hh')
data(HawaiiHymenoptera)
# For simplicity, we will select only those columns that we will use later
# Site_number: the site at which species were sampled
# LandUse: the land use at the sampled site
# Taxon_name_entered: the species sampled
# Measurement: the recorded abundance of the species
hh <- hh[,c('Site_number','LandUse','Taxon_name_entered','Measurement')]
# Make the site number a grouping variable (factor)
hh$Site_number <- factor(hh$Site_number)

# Subset to the dataset for just Pimpla punicipes
hh.sp <- droplevels(hh[(hh$Taxon_name_entered=="Pimpla punicipes"),])

# Load site-level data (loads as object 'hhs2'), from which we will obtain
# forest-cover estimates
data(HawaiiHymenopteraSitesSubset)

# The estimates of forest cover were not included in the species-level dataset,
# so we will match these from the site-level dataset
hh.sp$ForestCover <- hhs2$ForestCover[match(hh.sp$Site_number,hhs2$Site_number)]

# Remove sites that have missing (NA) forest-cover estimates
hh.sp <- droplevels(hh.sp[!is.na(hh.sp$ForestCover),])

str(hh.sp)
```

In the session on GLMs, we discovered that land use had a significant effect on the abundance of *P. punicipes*, but that the effect of forest cover was marginally non-significant. Here, we will investigate the same relatioonships, but using a multi-model rather than frequentist approach.

# Candidate Models

The first stage in a multi-model approach is to select a suitable set of candidate models. Ideally, this candidate set would consist of a series of models each reflecting some explanation (or hypothesis) for the way the system works. In reality, we often lack the information required to make an such an informed list of candidate models. One commonly used approach is to run models representing every possible combination of explanatory variables. Here, we will compare four different models of the abundance of *P. punicipes*:

1. A null (intercept-only) model
2. A model with only land use as an explanatory variable
3. A model with only forest cover
4. A model with both land use and forest cover

We know from the session on GLMs that the data describing the abundance of *P. punicipes* are over-dispersed, so we will start straight away by fitting negative binomial GLMs:

```{r,echo=TRUE,results=TRUE}
mNull <- glm.nb(Measurement~1,data=hh.sp)
mLU <- glm.nb(Measurement~LandUse,data=hh.sp)
mFC <- glm.nb(Measurement~ForestCover,data=hh.sp)
mCombined <- glm.nb(Measurement~LandUse+ForestCover,data=hh.sp)
```

# References