---
title: Working with Spatial Data in R
date: "`r Sys.Date()`"
author: "Tim Newbold"
output:
  rmdformats::downcute:
    code_folding: show
    self_contained: true
    thumbnails: false
    lightbox: true
---

# Overview

This session will cover the basics of working with spatial data in R. We will work mostly with the terra package, which is very versatile.

```{r LoadTerra, echo=TRUE, results=TRUE}
library(terra)
```

We will also use a number of other packages:

```{r LoadPackages, echo=TRUE, results=TRUE, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(tidyterra)
library(osmdata)
library(rnaturalearth)
library(geodata)
```

If you don't have any of the aforementioned packages already, you can install them using the _install.packages_ function.

# Types of Spatial Data

There are two major classes of spatial data that you will encounter: vector data, which consist of points, lines or polygons, or raster data, which describe values across a grid (typically with square grid cells, but sometimes hexagonal).

## Vector Data

There are lots of ways to load vector data in R. Ultimately, it is useful to convert vector data into the *SpatVector* data class used in the _terra_ package, as this package provides lots of functionality for working with and combining different types of spatial data. You can create a *SpatVector* object, or convert another spatial data type to this format, using the _vect_ function in _terra_

There are three main ways that we can load spatial data (and indeed all types of data) in R. First, there may be packages that allow us to read data automatically from a web location. Second, we can read data from a local file stored on your computer. Third, we can create a spatial data object from within R. We will cover all of these three methods, to some extent, during the course of this tutorial.

### Lines

Lines are a type of vector data describing linear features. For example, in ecology we might consider data on roads or hedgerows.

We will use the _osmdata_ package to query road data from Open Street Map. We will here obtain data on major roads (motorways, primary and secondary roads) in a region of New Zealand. This package reads in roads data in a bespoke format. We will first convert the data to a 'simple features' format using the _osmdata_sf_ function. Then, we can convert to a _terra_ package SpatVector format using the _vect_ function.

```{r GetRoadsData, echo=TRUE, results=TRUE}
# Define bounding box for roads query
opq(bbox = c(171.5,-43.6,172.9,-42.3)) %>%
  # Obtain data for major highways (motorways, primary and secondary roads)
  add_osm_feature(key = "highway", 
                  value = c("motorway", "primary", "secondary")) %>%
  # Convert to simple features class
  osmdata_sf() -> 
  nz_major

# Convert to terra SpatVector object (the osm_lines part of the original object
# contains the actual roads data we are interested in)
nz_major$osm_lines %>% vect() -> nz_major_sv
```
We will plot the roads map using ggplot and the geom_spatvector function from the _tidyterra_ package.

```{r PlotRoads, echo=TRUE, results=TRUE}
p <- ggplot() + 
  geom_spatvector(data = nz_major_sv) + 
  scale_x_continuous(name = "",limits = c(171.5,172.9)) + 
  scale_y_continuous(name = "",limits = c(-43.6,-42.3)) + 
  theme_classic()

p
```

Each line in a spatial vector lines dataset is described by a series of coordinates that form the vertices of each line. We can inspect these using the _geom_ function in _terra_:

```{r InspectRoadsData, echo=TRUE, results=TRUE}
# Show the geometry for the first road in the dataset
head(geom(nz_major_sv[1,]))
```

The first six rows here all refer to the same geometry, _(geom = 1)_, because the coordinates of these rows are all vertices of the first road in the dataset. In total, there are 3668 different roads (geometries) in this dataset.

### Polygons

Polygon data are very similar to vector line data, but the lines connect at the ends to form polygons.

Polygons are often used to describe the boundaries of regions. In ecological studies, we might be interested in administrative regions, such as countries, or ecological regions, such as biogeographic realms, biomes or ecoregions, or many other spatial regions.

Here, we will use the _rnaturalearth_ package to read in country borders, here the border of New Zealand. This package has the option to return the data directly as a _terra_ package 'SpatVector' object: _returnClass='sv'_.

```{r GetCountriesData, echo=TRUE, results=TRUE}
# Obtain polygon data describing the border of New Zealand
ne_countries(scale = 50,returnclass = 'sv',country = "New Zealand") -> nz
```

We can plot this polygon map using the same approach we used for the roads data.

```{r PlotCountries1, echo=TRUE, results=TRUE}
p <- ggplot() + 
  geom_spatvector(data = nz,fill="#5DA89955",col="#00000000") + 
  # Limit the span of the plot because one part of New Zealand (the Chatham
  # Islands) falls in the Western Hemisphere
  scale_x_continuous(name = "",limits = c(165,180)) + 
  scale_y_continuous(name = "",limits = c(-50,-30)) + 
  theme_classic()

p
```

We can then combine the plots for the polygons (New Zealand border) and lines (major roads).

```{r PlotCountries2, echo=TRUE, results=TRUE}
p <- ggplot() + 
  geom_spatvector(data = nz,fill="#5DA89955",col="#00000000") + 
  geom_spatvector(data = nz_major_sv,col="#2E2585") +
  # Limit the span of the plot because one part of New Zealand (the Chatham
  # Islands) falls in the Western Hemisphere
  scale_x_continuous(name = "",limits = c(171.5,172.9)) + 
  scale_y_continuous(name = "",limits = c(-43.6,-42.3)) + 
  theme_classic()

p
```

As with line data, polygons data are described a series of coordinates forming the vertices of each polygon, which we can see using the _geom_ function again.

```{r InspectCountriesData, echo=TRUE, results=TRUE}
head(geom(nz))
```

Here, all rows belong to the same geometry, _(geom = 1)_ (New Zealand), but the data are separated into 13 different polygons (parts). Here, the first six rows that we see are all vertices for the first polygon/'part'. With polygon data, the final column, _hole_ tells us whether a vertex belongs to an inner hole rather than the outer boundary (although there aren't any holes in the New Zealand map). 

### Points

The final type of vector data you will encounter describe single points in space.

To illustrate point data, we will use a dataset from within the PREDICTS database, which you will encounter later.

For this, you will need my _predictsFunctions_ package for working with the PREDICTS database. You can install this by running: _remotes::install_github("timnewbold/predicts-demo",subdir="predictsFunctions")_

```{r LoadPREDICTSPackage, echo=TRUE, results=TRUE}
library(predictsFunctions)
```

For now, don't worry about the details of this code. It just obtains the PREDICTS database, selects data from a single study on beetles in New Zealand, and calculates species richness for the locations sampled.

```{r GetPREDICTSData, echo=TRUE, results=TRUE, message=FALSE}
url("https://www.dropbox.com/scl/fi/0mqoacviqiiurrtplby2m/predicts.rds?rlkey=m5ijge7w1dthkvm5gbp652uel&dl=1") %>% readRDS() %>% droplevels() -> predicts_2016
url("https://www.dropbox.com/scl/fi/3dqv4ptkqkuswxx350xbx/predicts_2022.rds?rlkey=20g6v43l0vjdh3k8e7p533hri&dl=1") %>% readRDS() %>% droplevels() -> predicts_2022
full_join(predicts_2016,predicts_2022) -> predicts

ewers2007 <- droplevels(predicts[(predicts$SS=="CC1_2007__Ewers 1"),])

ewers2007 <- ewers2007 %>% CorrectSamplingEffort()

ewers2007 <- ewers2007 %>% MergeSites(silent=TRUE)

ewers2007 %>%
  droplevels() %>%
  SiteMetrics(extra.cols = c("Longitude","Latitude",
                             "Predominant_land_use",
                             "SSB","SSBS","Study_common_taxon","Country"),
              srEstimators = NULL) %>%
  mutate(LandUse = Predominant_land_use) %>% droplevels() -> sites

```

At the moment, we just have a data frame that contains the sampled locations, with their latitude and longitude coordinates. To work with these data as spatial data, we need to convert them into a proper spatial object. To do this, we can use the _vect_ function to create a vector spatial object from our data, specifying the 'Longitude' and 'Latitude' columns as the geometry (the spatial coordinates).

Because we are creating these data from scratch, we need to specify the coordinate and projection system. Much more on this later, but for now we will simply specify the coordinate system using the _crs_ function. We will use a simple geographical projection based on the World Geodetic System 1984 (WGS84) coordinate system (this is projection number 4326 in the EPSG list).

```{r MakePREDICTSSitesMap, echo=TRUE, eval=TRUE}

sitesMap <- vect(sites,geom=c('Longitude','Latitude'))
crs(sitesMap) <- "epsg:4326"

```

As before, we can plot the vector data using the _geom_spatvector_ function with _ggplot_. Again, we will add the map of New Zealand as background.

```{r PlotPREDICTSSites1, echo=TRUE, eval=TRUE}
p <- ggplot() + 
  geom_spatvector(data = nz,fill="#5DA89955",col="#00000000") + 
  geom_spatvector(data = sitesMap,size=1,col="#7E2954") + 
  scale_x_continuous(name = "",limits = c(165,180)) + 
  scale_y_continuous(name = "",limits = c(-50,-30)) + 
  theme_classic()

p
```

Zooming in to just the region where sampled locations lie, we can see the spatial arrangement of the points a bit better.

```{r PlotPREDICTSSites2, echo=TRUE, eval=TRUE}
p <- ggplot() + 
  geom_spatvector(data = nz,fill="#5DA89955",col="#00000000") + 
  geom_spatvector(data = sitesMap,size=1,col="#7E2954") + 
  scale_x_continuous(name = "",limits = c(172.3,172.5)) + 
  scale_y_continuous(name = "",limits = c(-42.65,-42.45)) + 
  theme_classic()

p
```

We could also add the roads data (here we only see one road becuase we are looking at a remote region in New Zealand)

```{r PlotPREDICTSSites3, echo=TRUE, eval=TRUE}
p <- ggplot() + 
  geom_spatvector(data = nz,fill="#5DA89955",col="#00000000") + 
  geom_spatvector(data = nz_major_sv,col="#2E2585") +
  geom_spatvector(data = sitesMap,size=1,col="#7E2954") + 
  scale_x_continuous(name = "",limits = c(172.3,172.5)) + 
  scale_y_continuous(name = "",limits = c(-42.65,-42.45)) + 
  theme_classic()

p
```

The full original dataset is still contained within the map object:

```{r InspectPREDICTSSitesData, echo=TRUE, eval=TRUE}
head(sitesMap)
```

Just as we did before with the other vector data types, we can inspect the geometry of the points map:

```{r InspectPREDICTSSitesMap, echo=TRUE, eval=TRUE}
head(geom(sitesMap))
```

Here, each row belongs to a different geometry (point) with specific (point) coordinates.

## Raster Data

We have now covered the three vector spatial data types. Now, we will move onto raster data, which have a different structure. In raster data, we have data values for each cell within a spatial grid. Generally this grid consists of square cells, but you may sometimes come across rasters with rectangular or even hexagonal grid cells.

Whereas before we read in vector spatial data using the _vect_ function in the *_terra_* package, now we will use the _rast_ function (also from the *_terra_* package) to read and create raster data.

We will first read in an example raster dataset from file. This dataset describes the proportion of natural habitat globally at a 30-arc-second resolution (see box, below), although for a lighter-weight download here we will work with a subset of these data just for New Zealand.

<div style="background-color: #f0f0f0; color: black; padding: 15px; border: none;">
  <strong style="font-size: 1.1em;">A Note on Spatial Resolution</strong><br>
  <p>For rasters that are in a geographic projection (see more on projection systems later), the spatial units of the  raster are degrees of longitude/latitude.</p>
  
  <p>Thus, the resolution of such a raster can be described in terms of degrees (1 degree, 0.5 degrees etc.). Or, it can be described in terms of arc-minutes. There are 60 arc-minutes in a degree, and so a 0.5 degree resolution is the same as 30 arc-minute resolution. Alternatively, the resolution can be given in arc-seconds. There are 60 arc-seconds in an arc-minute, and so 3600 arc-seconds in a degree.</p>
  
  <p>For a raster with square cells, you obviously need just one value to describe resolution. In the rare case of a raster with rectangular cells, you need two values.</p>
  
  <p>For a raster with hexagonal cells, resolution is described using a bespoke system. For example, in the H3 system (implemented in the _h3_ package, in case you are interested), resolutions  range from 0, which is a very coarse resolution where each hexagon covers approximately 4 million km<sup>2</sup>, to 15, which is an incredibly fine resolution, with each hexagon covering less than 1 m<sup>2</sup>.</p>
</div>

```{r ReadNaturalHabitatData, echo=TRUE, eval=TRUE}

rast("https://www.dropbox.com/scl/fi/y8jr7uawyp1c4zcrscgiy/ProportionNaturalHabitatNewZealand.tif?rlkey=e4ecfbapcaf934et3l93if7sb&dl=1") ->
  propn_natural

```
```{r PlotNaturalHabitat1, echo=TRUE, eval=TRUE}
p <- ggplot() + 
  geom_spatraster(data = propn_natural) + 
  scale_fill_grass_c(palette = 'forest_cover') + 
  theme_classic()

p
```

```{r, echo=TRUE, eval=TRUE}
p <- ggplot() + 
  geom_spatraster(data = propn_natural) + 
  scale_fill_grass_c(palette = 'forest_cover') + 
  geom_spatvector(data = nz_major_sv,col="#2E2585") +
  geom_spatvector(data = sitesMap,size=1,col="#7E2954") + 
  scale_x_continuous(name = "",limits = c(172.3,172.5)) + 
  scale_y_continuous(name = "",limits = c(-42.66,-42.45)) + 
  theme_classic()

p
```

Similar to a vector object, where each point or line/polygon vertex is described by a pair of coordinates (longitude and latitude), the raster object contains a pair of coordinates representing the centre of each cell. We can show the coordinates using the _crds_ function.

```{r RasterCoordinates, echo=TRUE, eval=TRUE}
head(crds(propn_natural))
```

Besides the coordinates for each grid cell, the other main information in a raster object are the values of the variable shown in the raster (here the proportional coverage of natural habitats). We can show these values using the _values_ function.

```{r RasterValues, echo=TRUE, eval=TRUE}
summary(values(propn_natural))
```

```{r RasterValuesHistogram, echo=TRUE, eval=TRUE, warning=FALSE}
ggplot() + 
  geom_histogram(mapping = aes(x = values(propn_natural))) + 
  theme_classic()
```

It is very simple to perform arithmetic operations on raster objects in R. You simply pass the whole raster into a calculation just as you would any other object. For example, let's convert our raster describing proportional coverage of natural habitats into percentage values by multiplying by 100:

```{r PercentNaturalHabitat, echo=TRUE, eval=TRUE, results=TRUE}
perc_natural <- propn_natural * 100

# We have to use the basic colour scheme here, as the one we used before
# only works on 0 to 1 values
p <- ggplot() + 
  geom_spatraster(data = perc_natural) + 
  theme_classic()

p
```

You can also combine two or more separate rasters into a calculation (as long as they have an identical spatial domain and resolution).

# Coordinate and Projection Systems

All maps captured as 2-dimensional objects are a representation of a 3-dimensional (spherical) world. The way that this conversion works depends on the coordinate and projection system used.

The simplest way of representing spatial data is using a geographical projection, where space is described in terms of degrees of longitude and latitude, and the spatial unit is degrees.

Because the Earth isn't perfectly spherical, longitude and latitude coordinates are mapped using a theoretical representation of the Earth's surface. This is called the 'geodetic' system. Different systems can produce coordinates that vary, sometimes by as much as several hundred metres. A very commonly used coordinate system is the World Geodetic System 1984 (WGS84 for short).

For this part of the tutorial, we will work with global climate data from WorldClim, which we can obtain using the _geodata_ package in R.

```{r GetTemperatureData, echo=TRUE, eval=TRUE, results=TRUE}
# Read average temperature at 5 arc-minutes resolution
temp <- worldclim_global(var = "tavg",res = 5,path = tempdir())
```

These data describe the average temperature in each of the months of the year, at a resolution of 5 arc-minutes, and mapped with a geographical projection and the WGS84 coordinate system.

We will first average the estimates of temperature across the 12 months, to obtain a single annual average temperature.

```{r AverageTemperatureData, echo=TRUE, eval=TRUE, results=TRUE}
AnnAvgTemp <- mean(temp)
```

```{r PlotTemperatureWGS84, echo=TRUE, eval=TRUE, results=TRUE}
p <- ggplot() + 
  geom_spatraster(data = AnnAvgTemp) + 
  scale_fill_grass_c(palette = 'celsius') + 
  theme_classic()

p
```

```{r, echo=TRUE, eval=TRUE, results=TRUE}
summary(cellSize(AnnAvgTemp))
```
# Combining Spatial Data and Making Calculations

# Exercises
